Data Specialization for a Staged Lambda Calculus

### 2015-02-28 paper/examples.tex
hyphenating "first-stage" and "second-stage" when appropriate

Pipeline programs can be directly expressed as staged functions in \lang\, with splitting 
used to generate monostaged code required for each GPU pipeline stage
(as well as the plumbing to communicate values between the stages).
Composition of such functions in \lang\ allows more complex pipelines to be expressed in a modular programming style
that is becoming increasingly desirable in graphics shading languages\,\cite{Foley:2011,He:2014}. A simple example of composition of staged functions (representing GPU pipeline programs) is given below.  The code defines a pipeline as a staged function that takes an object definition in first stage (\texttt{obj}) and a pixel coordinate (\texttt{xy}) in the second stage and emits the color of the object at the specified pixel. The code defines a pipeline combinator \texttt{shade} which modulates the results of the object's reflectance at a given pixel with the results of a function that computes the object's albedo (surface color).  
%
\begin{lstlisting} 
1`datatype object = ...`
2`atsignnext{
  type coord = int * int
  type color = int
}`
1`type pipeline = object * $`2`coord`1` -> `$2`color`

>>>>

Pipeline programs can be directly expressed as staged functions in \lang\, with splitting 
used to generate monostaged code required for each GPU pipeline stage
(as well as the plumbing to communicate values between the stages).
Composition of such functions in \lang\ allows more complex pipelines to be expressed in a modular programming style
that is becoming increasingly desirable in graphics shading languages\,\cite{Foley:2011,He:2014}. A simple example of composition of staged functions (representing GPU pipeline programs) is given below.  The code defines a pipeline as a staged function that takes an object definition in the first stage (\texttt{obj}) and a pixel coordinate (\texttt{xy}) in the second stage and emits the color of the object at the specified pixel. The code defines a pipeline combinator \texttt{shade} which modulates the results of the object's reflectance at a given pixel with the results of a function that computes the object's albedo (surface color).  
%
\begin{lstlisting} 
1`datatype object = ...`
2`atsignnext{
  type coord = int * int
  type color = int
}`
1`type pipeline = object * $`2`coord`1` -> `$2`color`

### 2015-02-28 paper/overview.tex
hyphenating "first-stage" and "second-stage" when appropriate

As the splitting algorithm scans \texttt{qss} for computations that
can be performed in the first stage, it also collects computations
that must be left to the second stage in a separate function
\texttt{qs2}.  This function, which is executed in the second pass,
takes as argument the boundary and the second-stage argument (the
rank) and performs a lookup in the boundary data structure.  But how
does \texttt{qs2} knows to traverse the tree? It does not. Since
the splitting algorithm recorded the control flow of the first stage
execution in the boundary data structure, \texttt{qs2} simply follows
this control structure and performs at each point the parts of the
computation from \texttt{qss} that can now be performed in the
presence of the second-stage argument (the rank).  In the context of
our example, this performs a lookup on the boundary data structure by
using the supplied rank.

>>>>

As the splitting algorithm scans \texttt{qss} for computations that
can be performed in the first stage, it also collects computations
that must be left to the second stage in a separate function
\texttt{qs2}.  This function, which is executed in the second pass,
takes as argument the boundary and the second-stage argument (the
rank) and performs a lookup in the boundary data structure.  But how
does \texttt{qs2} knows to traverse the tree? It does not. Since
the splitting algorithm recorded the control flow of the first-stage
execution in the boundary data structure, \texttt{qs2} simply follows
this control structure and performs at each point the parts of the
computation from \texttt{qss} that can now be performed in the
presence of the second-stage argument (the rank).  In the context of
our example, this performs a lookup on the boundary data structure by
using the supplied rank.

### 2015-02-28 paper/relatedwork.tex
hyphenating "first-stage" and "second-stage" when appropriate

\paragraph{Frequency-reduction via pass separation.}
Perhaps the closest work to ours is the algorithm for ``data
specialization'' given by Knoblock and Ruf~\cite{knoblock96}.  Like
our work, Knoblock and Ruf are concerned about the problem of
statically splitting an explicitly staged program into two stages.
However, they only consider simple first-order language and
straight-line programs and also treat all computations guarded by
second stage conditionals as second stage computations.

>>>>

\paragraph{Frequency-reduction via pass separation.}
Perhaps the closest work to ours is the algorithm for ``data
specialization'' given by Knoblock and Ruf~\cite{knoblock96}.  Like
our work, Knoblock and Ruf are concerned about the problem of
statically splitting an explicitly staged program into two stages.
However, they only consider simple first-order language and
straight-line programs and also treat all computations guarded by
second-stage conditionals as second-stage computations.

### 2015-02-28 paper/splitting.tex
hyphenating "first-stage" and "second-stage" when appropriate

Splitting statically transforms terms in \lang\ into a pair of
monostaged terms ($f_1$ and $f_2(-)$) that correspond to the first
stage and second stage computations in the original term. Since full
evaluation of world \bbtwo\ terms (both first stage and second stage
evaluation) only produces values in stage two, the idea of splitting
is to generate terms where evaluating $f_1$ in a first pass, and then
evaluating $f_2$ on the result of the first pass, yields the same
value as full evaluation of the original \lang\ program.

Similarly, when splitting \bbonem\ terms, first pass evaluation of
$f_1$ should produce the same value as first stage evaluation of the
original \lang\ term.  However, complexity arises because
\bbonem\ terms may also contain computations to be evaluated in the
second stage. (For example, a \bbonem\ term of type $\fut \rmint$
results in an \rmint\ value in the second stage.) Thus we also want
the second pass evaluation of $f_2$ on the outputs of $f_1$ to yield
the same value as fully evaluating all second stage computations in
the original \bbonem\ term. Finally, when splitting \bbonep\ terms,
which only produce values in the first stage, we only need to compare
the result of $f_1$ with that of the \lang\ term.

Handling the asymmetry between the results of evaluating
\bbonem\ terms (which may produce results at both the first and second
stage) and \bbtwo\ terms (which only produce second stage results) is
fundamental to the structure of the splitting algorithm presented in
this section.

\textbf{\bbtwo\ term splitting.}  For any $\coltwo e A$ which reduces
to a residual $q$ (via $\diatwo e q$) and then a final value $v$ (via
$\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
yields a pair of monostaged terms $\pipeS p l r$. ($p$ and $l.r$
contain all first and second stage computations in $e$ respectively.)
Executing $p$ in a first pass produces $b$ (via $\reduce p b$). Then,
in the second pass, $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
$v$.  Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

@@

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
relation~$\splitonesym$) again yields the pair of monostage terms
$\pipeM c l r$. Evaluating $c$ in the first pass produces the pair
$(i,b)$, where the first element $i$ matches the first stage results
contained in $v$.  For example, in the simple case where $\rtab \xi v$
contains entirely first stage computation, such as:
\begin{lstlisting}
2`[]`1` 3`gr{`1`literalone+literaltwo`3`}`
\end{lstlisting}
the first stage result is the value 3.  The second element of the pair
is the boundary value used as input to the resumer in second pass
evaluation as $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
result produced by second stage computations in $e$.  For example, in
the case where $\rtab \xi v$ contains entirely second stage
computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into their first and second stage results. We then
describe the algorithm for splitting both \bbonem\ terms
($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).

@@

%Note that since $\coltwo e A$ reduces to an entirely second-stage
%residual, its first-stage computations only exist to generate input
%for second stage computations.  In contrast, since $\colmix e A$
%reduces to a multi-stage residual table and partial value, its first
%stage computations serve to both generate the value used as input for
%the second stage {\em and also} produce a first-stage value.
%Correspondingly, the precomputation $p$ resulting from splitting
%\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
%the resumer; but the combined term $c$ resulting from splitting
%\bbonem\ terms reduces to a tuple containing both the first stage {\em
%  immediate result} $i$ and the boundary value $b$ (motivating the
%name {\em combined term}).

@@

%Masking separates residual tables and partial values into first stage
%results and second stage residuals.

Masking (given by the relation $\vsplito$) converts a residual table $\xi$
and associated partial value $v$ into a masked value $\mval i q$.
%that explicitly separates the first stage components of 
%result from the residual.
%where $i$ is a monostage value encoding all
%first stage components of the partial value, and $q$ is a monostage
%term encoding all of the second-stage computations of the partial
%value.
The first part of the masked value, $i$, is a monostage value holding
the result of all first stage components of $v$.  The second
component, $q$, is a residual representing all second stage
computations in $\xi$ and $v$.

@@

%Masking is not a transformation performed as part of the splitting
%algorithm.  Instead, masking defines an intuitive notion of the first
%stage and second stage ``components'' of residual tables and partial
%values, and it is used to define the requirements of what output
%splitting of \bbonem\ terms must produce.
To provide intuition about the intent of masking, consider the
following residual table and partial value:
\begin{lstlisting}
2`[yhat |-> 1+2]`1` (next{`2`yhat`1`}, injL (3`gr{`1`7`3`}`, next{`2`yhat`1`}))`
\end{lstlisting}
To construct the value $i$ representing its first-stage components,
masking first redacts all second stage (blue) parts, along with the
surrounding \texttt{next} annotations. (This redaction removes the
entire residual table since it only represents second stage
computations.) The resulting ``holes'' in the term are replaced with
unit values.
%\begin{lstlisting}
%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
%\end{lstlisting}
%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
\begin{lstlisting}
1`((), injL (3`gr{`1`7`3`}`,()))`
\end{lstlisting}
Finally, masking drops \texttt{gr} annotations yielding:
\begin{lstlisting}
1`((), injL (7,()))`
\end{lstlisting}
To construct the residual $q$ (corresponding to second stage computations) masking first redacts all \texttt{gr} blocks
(replacing them with unit), \texttt{next} annotations, and injection tags:
%\begin{lstlisting}
%2`[yhat |-> 1+2]`1` (`4`####`2`yhat`4`#`1`, `4`####`1` (`4`####`1`,`4`####`2`yhat`4`#`1`))`
%\end{lstlisting}
%Where the \texttt{gr} blocks left a hole, we put in a unit values:
\begin{lstlisting}
2`[yhat |-> 1+2] (yhat, ((),yhat))`
\end{lstlisting}
Then the residual table is reified into \texttt{let} bindings,
yielding:
\begin{lstlisting}
2`let yhat = 1+2 in (yhat, ((),yhat))`
\end{lstlisting}
A precise definition of the masking relation is given in
\ref{fig:valMask}.

%Since lambdas may represent multi-stage computations, masking splits
%the body of lambdas as general world \bbonem\ terms (as described in
%\ref{sec:split-one}), and packages the resulting terms as functions in
%both the first and second stage parts of the masked value.

@@

In other words, $c'$ cases on $i_1$ to run the first branch of first stage
code. The boundary value for that branch, $b_2$, is packaged as $\tup {b_1, \inl
{b_2}}$. The resumer of the predicate uses $b_1$, and $\inl{b_2}$ signals to run
the first branch's resumer. The second branches of $c'$ and $l'.r'$ are
similarly constrained by the $\texttt{inr}$ case.

@@

\paragraph{Next and Prev}
Since \texttt{next} and \texttt{prev} do nothing more than convert between 
second stage terms at \bbtwo\ and second stage terms encapsulated within $\fut$s at \bbonem,
their splitting rules are relatively simple.
Given an $\splittwo e ? p l r$ and $\diatwo e q$, the splitting invariant says,
\[
	\reduce p b \text{ and } [b/l]r \equiv q
\]
The splitting invariant for $\splitone {\next {e'}} ? {c'} {l'.r'}$,
where $\diaone {\next {e'}} {{\hat y} \mapsto q} {\next {\hat y}}$
and $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$
then requires
\[
	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin {\hat y} q {\hat y}
\]
which can be fulfilled using $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.

@@

The splitting algorithm described in the previous subsections operates
purely on the local structure of \lang\ terms.  One artifact of this design
is that splitting \bbonem\ terms may generate resumers containing
unnecessary logic.  For example, the rule for splitting
\bbonem\ \texttt{case} terms inserts the tag from the \texttt{case}
argument into the boundary value, then decodes this tag in the
resumer. This logic occurs regardless of whether the terms forming the
branches of the \texttt{case} contain second stage computations.
Worse, if this \texttt{case} appeared in the body of a recursive
function with no other second stage computations, splitting would
generate a resumer with (useless) recursive calls.

Rather than attempt global optimization of the outputs of splitting,
we instead leverage the type system to indicate when a term contains
no second-stage computations by adding a third world \bbonep\ whose
terms are purely first stage.  While not necessary for ordinary staged
term evaluation (\ref{sec:stagedsemantics}), this annotation is
particularly useful for splitting. Specifically, splitting $\pure e$
terms trivially yields a unit resumer.

The \texttt{partition} function in the \texttt{quickselect} example in
\ref{sec:exampleQS} is an example of such a function that contains no
second stage computations, but, if not defined within a $\pure$
annotation, would cause splitting to generate a recursive resumer.

>>>>

Splitting statically transforms terms in \lang\ into a pair of
monostaged terms ($f_1$ and $f_2(-)$) that correspond to the first-stage 
and second-stage computations in the original term. Since full
evaluation of world \bbtwo\ terms (both first-stage and second-stage
evaluation) only produces values in stage two, the idea of splitting
is to generate terms where evaluating $f_1$ in a first pass, and then
evaluating $f_2$ on the result of the first pass, yields the same
value as full evaluation of the original \lang\ program.

Similarly, when splitting \bbonem\ terms, first pass evaluation of
$f_1$ should produce the same value as first-stage evaluation of the
original \lang\ term.  However, complexity arises because
\bbonem\ terms may also contain computations to be evaluated in the
second stage. (For example, a \bbonem\ term of type $\fut \rmint$
results in an \rmint\ value in the second stage.) Thus we also want
the second pass evaluation of $f_2$ on the outputs of $f_1$ to yield
the same value as fully evaluating all second-stage computations in
the original \bbonem\ term. Finally, when splitting \bbonep\ terms,
which only produce values in the first stage, we only need to compare
the result of $f_1$ with that of the \lang\ term.

Handling the asymmetry between the results of evaluating
\bbonem\ terms (which may produce results at both the first and second
stage) and \bbtwo\ terms (which only produce second-stage results) is
fundamental to the structure of the splitting algorithm presented in
this section.

\textbf{\bbtwo\ term splitting.}  For any $\coltwo e A$ which reduces
to a residual $q$ (via $\diatwo e q$) and then a final value $v$ (via
$\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
yields a pair of monostaged terms $\pipeS p l r$. ($p$ and $l.r$
contain all first- and second-stage computations in $e$ respectively.)
Executing $p$ in a first pass produces $b$ (via $\reduce p b$). Then,
in the second pass, $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
$v$.  Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

@@

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
relation~$\splitonesym$) again yields the pair of monostage terms
$\pipeM c l r$. Evaluating $c$ in the first pass produces the pair
$(i,b)$, where the first element $i$ matches the first-stage results
contained in $v$.  For example, in the simple case where $\rtab \xi v$
contains entirely first-stage computation, such as:
\begin{lstlisting}
2`[]`1` 3`gr{`1`literalone+literaltwo`3`}`
\end{lstlisting}
the first-stage result is the value 3.  The second element of the pair
is the boundary value used as input to the resumer in second pass
evaluation as $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
result produced by second-stage computations in $e$.  For example, in
the case where $\rtab \xi v$ contains entirely second-stage
computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second-stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into their first- and second-stage results. We then
describe the algorithm for splitting both \bbonem\ terms
($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).

@@

%Note that since $\coltwo e A$ reduces to an entirely second-stage
%residual, its first-stage computations only exist to generate input
%for second-stage computations.  In contrast, since $\colmix e A$
%reduces to a multi-stage residual table and partial value, its first
%stage computations serve to both generate the value used as input for
%the second stage {\em and also} produce a first-stage value.
%Correspondingly, the precomputation $p$ resulting from splitting
%\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
%the resumer; but the combined term $c$ resulting from splitting
%\bbonem\ terms reduces to a tuple containing both the first stage {\em
%  immediate result} $i$ and the boundary value $b$ (motivating the
%name {\em combined term}).

@@

%Masking separates residual tables and partial values into first-stage
%results and second stage residuals.

Masking (given by the relation $\vsplito$) converts a residual table $\xi$
and associated partial value $v$ into a masked value $\mval i q$.
%that explicitly separates the first stage components of 
%result from the residual.
%where $i$ is a monostage value encoding all
%first-stage components of the partial value, and $q$ is a monostage
%term encoding all of the second-stage computations of the partial
%value.
The first part of the masked value, $i$, is a monostage value holding
the result of all first-stage components of $v$.  The second
component, $q$, is a residual representing all second-stage
computations in $\xi$ and $v$.

@@

%Masking is not a transformation performed as part of the splitting
%algorithm.  Instead, masking defines an intuitive notion of the first-stage 
%and second-stage ``components'' of residual tables and partial
%values, and it is used to define the requirements of what output
%splitting of \bbonem\ terms must produce.
To provide intuition about the intent of masking, consider the
following residual table and partial value:
\begin{lstlisting}
2`[yhat |-> 1+2]`1` (next{`2`yhat`1`}, injL (3`gr{`1`7`3`}`, next{`2`yhat`1`}))`
\end{lstlisting}
To construct the value $i$ representing its first-stage components,
masking first redacts all second-stage (blue) parts, along with the
surrounding \texttt{next} annotations. (This redaction removes the
entire residual table since it only represents second-stage
computations.) The resulting ``holes'' in the term are replaced with
unit values.
%\begin{lstlisting}
%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
%\end{lstlisting}
%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
\begin{lstlisting}
1`((), injL (3`gr{`1`7`3`}`,()))`
\end{lstlisting}
Finally, masking drops \texttt{gr} annotations yielding:
\begin{lstlisting}
1`((), injL (7,()))`
\end{lstlisting}
To construct the residual $q$ (corresponding to second-stage computations) masking first redacts all \texttt{gr} blocks
(replacing them with unit), \texttt{next} annotations, and injection tags:
%\begin{lstlisting}
%2`[yhat |-> 1+2]`1` (`4`####`2`yhat`4`#`1`, `4`####`1` (`4`####`1`,`4`####`2`yhat`4`#`1`))`
%\end{lstlisting}
%Where the \texttt{gr} blocks left a hole, we put in a unit values:
\begin{lstlisting}
2`[yhat |-> 1+2] (yhat, ((),yhat))`
\end{lstlisting}
Then the residual table is reified into \texttt{let} bindings,
yielding:
\begin{lstlisting}
2`let yhat = 1+2 in (yhat, ((),yhat))`
\end{lstlisting}
A precise definition of the masking relation is given in
\ref{fig:valMask}.

%Since lambdas may represent multi-stage computations, masking splits
%the body of lambdas as general world \bbonem\ terms (as described in
%\ref{sec:split-one}), and packages the resulting terms as functions in
%both the first- and second-stage parts of the masked value.

@@

In other words, $c'$ cases on $i_1$ to run the first branch of first-stage
code. The boundary value for that branch, $b_2$, is packaged as $\tup {b_1, \inl
{b_2}}$. The resumer of the predicate uses $b_1$, and $\inl{b_2}$ signals to run
the first branch's resumer. The second branches of $c'$ and $l'.r'$ are
similarly constrained by the $\texttt{inr}$ case.

@@

\paragraph{Next and Prev}
Since \texttt{next} and \texttt{prev} do nothing more than convert between 
second-stage terms at \bbtwo\ and second-stage terms encapsulated within $\fut$s at \bbonem,
their splitting rules are relatively simple.
Given an $\splittwo e ? p l r$ and $\diatwo e q$, the splitting invariant says,
\[
	\reduce p b \text{ and } [b/l]r \equiv q
\]
The splitting invariant for $\splitone {\next {e'}} ? {c'} {l'.r'}$,
where $\diaone {\next {e'}} {{\hat y} \mapsto q} {\next {\hat y}}$
and $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$
then requires
\[
	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin {\hat y} q {\hat y}
\]
which can be fulfilled using $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.

@@

The splitting algorithm described in the previous subsections operates
purely on the local structure of \lang\ terms.  One artifact of this design
is that splitting \bbonem\ terms may generate resumers containing
unnecessary logic.  For example, the rule for splitting
\bbonem\ \texttt{case} terms inserts the tag from the \texttt{case}
argument into the boundary value, then decodes this tag in the
resumer. This logic occurs regardless of whether the terms forming the
branches of the \texttt{case} contain second-stage computations.
Worse, if this \texttt{case} appeared in the body of a recursive
function with no other second-stage computations, splitting would
generate a resumer with (useless) recursive calls.

Rather than attempt global optimization of the outputs of splitting,
we instead leverage the type system to indicate when a term contains
no second-stage computations by adding a third world \bbonep\ whose
terms are purely first-stage.  While not necessary for ordinary staged
term evaluation (\ref{sec:stagedsemantics}), this annotation is
particularly useful for splitting. Specifically, splitting $\pure e$
terms trivially yields a unit resumer.

The \texttt{partition} function in the \texttt{quickselect} example in
\ref{sec:exampleQS} is an example of such a function that contains no
second-stage computations, but, if not defined within a $\pure$
annotation, would cause splitting to generate a recursive resumer.
